\addchap{Abbreviations}
% \addchap{Abbreviations and symbols} 
\begin{tabularx}{.45\textwidth}{lX} 
1, 2, 3 & 1st, 2nd, 3rd person\\
\textsc{acl} & accidental\\
\textsc{adjct} & adjunct\\
\textsc{adnom} & adnominal\\
AdPoss. & adnominal possessive\\
\textsc{adv} & ad{verb}\\
\textsc{adv.t} & ad{verb}, temporal\\
\textsc{advs} & {adversative}\\
\textsc{affr} & affricate\\
\textsc{affx} & {affixation}\\
\textsc{ag} & agent\\
\textsc{agt} & agentive\\
\textsc{al} & alienable\\
Alt. & alternative\\
\textsc{alv} & alveolar\\
\textsc{an} & animate\\
\textsc{apr} & approximant\\
\textsc{a-p-ult} & antepenultimate\\
\textsc{argt} & argument\\
\textsc{assct} & association\\
\textsc{attr} & attributive\\
\textsc{aug} & augmentation\\
\textsc{ben} & beneficiary\\
\textsc{bw} & base word\\
\textsc{c} & consonant\\
\textsc{caus} & {causative}\\
\textsc{circ} & circumstance\\
\textsc{cl} & clause\\
\textsc{cmpr} & comparative\\
\textsc{cnj} & {conjunction}\\
\textsc{cst} & contrastive\\
\textsc{d.dist} & {demonstrative}, distal\\
\textsc{d.prox} & {demonstrative}, proximal\\
\textsc{dem} & {demonstrative}\\
\textsc{dim} & diminution\\
\textsc{diph} & diphthong\\
\end{tabularx} 
\begin{tabularx}{.45\textwidth}{lX} 
\textsc{do} & direct object\\
\textsc{dy} & dynamic\\
\textsc{edc} & education\\
\textsc{+edc} & better educated\\
\textsc{-edc} & less educated\\
\textsc{els} & {elision}\\
\textsc{emph} & emphasis, emphatic\\
\textsc{fric} & fricative\\
\textsc{glot} & glottal\\
\textsc{hum} & human\\
\textsc{ilct} & interlocutor\\
\textsc{inal} & inalienable\\
\textsc{inan} & inanimate\\
\textsc{ins} & instrument\\
\textsc{int} & {interrogative}\\
\textsc{intens} & {intensity}\\
\textsc{intr} & intransitive\\
Is & {Isirawa}\\
k.o. & kind of\\
\textsc{l.dist} & {locative}, distal\\
\textsc{l.med} & {locative}, medial\\
\textsc{l.prox} & {locative}, proximal\\
\textsc{lab} & labial\\
\textsc{lat-aprx} & lateral-approximant\\
\textsc{lig} & ligature\\
lim. & limited\\
\textsc{liq} & liquid\\
\textsc{loc} & {locative}\\
\textsc{loct} & location\\
marg. & marginal\\
med. & medium\\
\textsc{mod} & modifier\\
\textsc{n} & {noun}\\
\textsc{n.com} & {noun}, common\\
\textsc{n.loc} & {noun}, location\\
\textsc{n.time} & {noun}, time-denoting\\
\end{tabularx}
 

\begin{tabularx}{.45\textwidth}{lX} 
\textsc{nas} & nasal\\
\textsc{neg} & {negation}, negative\\
\textsc{neg.imp} & negative {imperative}\\
\textsc{nmlz} & nominalizer\\
\textsc{nom} & nominal\\
\textsc{np} & {noun} phrase\\
\textsc{num} & {numeral}\\
\textsc{o} & object\\
Obl. & oblique\\
\textsc{obstr} & obstruent\\
\textsc{omv} & object of mental {verb}\\
\textsc{orth} & orthography\\
Orthogr. & orthography\\
oSb & older sibling\\
\textsc{outsd} & outsider\\
\textsc{pal} & palatal\\
\textsc{pal-alv} & palato-alveolar\\
\textsc{pat} & patient\\
\textsc{pfx} & prefix\\
\textsc{phon} & phoneme\\
\textsc{pl} & plural\\
\textsc{pl-hold} & {placeholder}\\
\textsc{plos} & plosive\\
\textsc{pn} & {proper noun}\\
\textsc{pol} & politics\\
\textsc{poss} & possessive\\
\textsc{possm} & possessum\\
\textsc{possr} & possessor\\
\textsc{pp} & {prepositional phrase}\\
prec. & precedes\\
\textsc{pred} & predicate, predicative\\
\textsc{pro} & personal {pronoun}\\
\textsc{prod} & productivity\\
\textsc{pronom} & pronominal\\
\textsc{p-ult} & penultimate\\
\textsc{qt} & {quantifier}\\
\textsc{quant} & quantity\\
\textsc{rc} & relative clause\\
\textsc{rdp} & reduplicant\\
\textsc{rec} & recipient\\
\end{tabularx} 
\begin{tabularx}{.45\textwidth}{lX} 
\textsc{recp} & reciprocal\\
\textsc{rel} & religion\\
\textsc{rel} & {relativizer}\\
\textsc{res} & result\\
\textsc{ret} & retention\\
\textsc{rhot} & rhotic\\
\textsc{s} & subject\\
s.o. & someone\\
s.th. & something\\
\textsc{sg} & singular\\
SI & {Standard Indonesian}\\
Sim. & {similarity}\\
\textsc{spk} & speaker\\
\textsc{spm} & speech mistake\\
\textsc{st} & stative\\
\textsc{stat} & status\\
\textsc{+stat} & higher social status\\
\textsc{-stat} & lower social status\\
\textsc{std} & standard of comparison\\
\textsc{supl} & superlative\\
\textsc{sylb} & syllable\\
\textsc{top} & topic\\
\textsc{tru} & truncated\\
\textsc{ult} & ultimate\\
\textsc{up} & unclear pronunciation\\
\textsc{uv} & undergoer voice\\
\textsc{v} & vowel\\
\textsc{v} & {verb}\\
\textsc{v.bi} & {verb}, {bivalent}\\
\textsc{v.mo} & {verb}, {monovalent}\\
\textsc{v.tri} & {verb}, {trivalent}\\
\textsc{vblz} & verbalizer\\
\textsc{vdi} & {verb}, ditransitive\\
\textsc{vel} & velar\\
\textsc{voc} & vocative\\ 
\textsc{vp} & {verb} phrase\\
\textsc{vsi} & {verb}, intransitive stative\\
\textsc{vtr} & {verb}, transitive\\
ySb & younger sibling\\
\\
\end{tabularx} 