\addchap{Acknowledgments}
\begin{refsection}

%content goes here

The dissertation and this book could not have been written without the help of many people who have assisted, supported, guided, and encouraged me over the last years.

First of all I want to thank Kori\textsuperscript{†} and Sarlota\textsuperscript{†} Merne and their large family in \ili{Sarmi} and Webro who welcomed me into their house and family. They were always willing to help me, patiently answered all my questions, and let me record their conversations. Their love and friendship made me feel at home during my stays in \ili{Sarmi}. I can only mention of few other members of their family and guests here: Absalom, Alex, Anto, Aris, Beni, Desi, Dina, Domi, Edo, Elda, Febe, Fiki, Fritz\textsuperscript{†}, Hosea, Ice, Ici, Lina, Linda, Lodo, Martinus, Milka, Nela\textsuperscript{†}, Nesti, Ponti, Rolant, Rut, Salome, Sara, Sela, Siska, Soko, Thomas, Yako, Yos, Yulius, and Yusup.

Next I want to thank my friends and my two home churches in Hoffnungsthal and Oberhausen who supported me financially, prayed for me, and encouraged me throughout this research project. I would especially like to thank my home Bible study group in Hoffnungsthal. I am also very grateful for the scholarships which the Ken Pike Endowment Fund and SIL Asia Area provided over the past few years. Another word of thanks goes to my office in Germany with its wonderful staff. I am so grateful how my supervisors, colleagues, and friends encouraged and assisted me in this entire undertaking. I thank you all for standing beside me so faithfully

A word of thanks also goes to Chip Sanders of SIL International and to Mark Donohue and the participants of ISMIL 11 in Manokwari who encouraged me to take on this research project.

I want to express my gratitude toward my promotor Maarten Mous and my co-pro\-mo\-tor Marian Klamer. I want to thank Maarten for his constant support and encouragement during the process of this research project. I am very grateful to Marian for her supervision and for her commitment and encouragement throughout the years. I also want to thank her for her precise and detailed comments on different drafts of my original dissertation, which greatly improved its quality.

I owe deep gratitude to my Papuan consultant and friend Ben Rumaropen. Working with him on Papuan Malay not only deepened my understanding of the language but was also a great joy and pleasure. Kawang, trima-kasi e?!

I am very grateful to Antoinette Schapper for her generous support throughout the years and for working with me on different chapters of my dissertation. I also thank René van den Berg who read the first draft of my very first chapter so many years ago. His encouragement back then and throughout the past few years made a big difference for me in the entire process of this project.

\newpage  
\largerpage
A number of scholars took the time to read individual chapters of my original dissertation. Thank you Geert Booij, Lourens de Vries, Harald Hammarström, Tom Headland, Erin Hesse, Paul Kroeger, Leonid Kulikov, Paul Lewis, Francesca Moro, Rick Nivens, Tom Payne, Christian Rapold, Uri Tadmor, Johnny Tjia, and Wilco van den Heuvel. Another word of thanks goes to those scholars and colleagues who discussed various aspects related to Papuan Malay with me, and assisted me in many different ways. I can only mention a few here: Jermy Balukh, Ted Bergman, Robert Blust, John Bowden, Joyce Briley, Adrian Clynes, Chuck Grimes, Arne Kirchner, Maarten Kossmann, Oliver Kroeger, Waruno Mahdi, Ver\-onika Mattes, David Mead, Roger Mills, Edith Moravcsik, Scott Paauw, Angela Prinz, Werner Riderer, Jim Roberts, John Roberts, Silke Sauer, Ellie Scott, Graham Scott, Oliver Stegen, Joyce Sterner, Irene Tucker, Aone van Engelenhoven, and Catharina Williams-van Klinken.

A word of thank also goes to the participants of the AN-Lang (\ili{Austronesian} Languages) Mailing list and the SIL Linguistics Mailing list for responding to my inquiries and helping me in getting a better grasp of Papuan Malay.

I also want to thank the two anonymous reviewers, LangSci’s consulting editor, and Martin Haspelmath, one of LangSci’s general editors. Their comments, suggestions and corrections greatly helped to improve the quality of this book.

I am very grateful to the staff of SIL Asia Area, SIL Indonesia, and YABN, particularly Jackie Menanti, who provided me generously with their help in so many different ways. Another word of thanks goes to Emma Onim for transcribing some of the recorded texts, and to Lodowik Aweta for participating in the recording of the Papuan Malay \isi{word list}. Thanks also go to Allan and Karen Buseman and Marcus Jaeger for their help with Toolbox and the Citavi.com staff for their wonderful bibliography and citation program.

I owe deep gratitude to my colleagues Lenice and Jeff Harms. Staying with them in \ili{Sentani}, and being able to share my fascination with Papuan Malay with Lenice was a joy and pleasure. I am also grateful to Rita Eltgroth and Joyce Briley. I enjoyed sharing offices in the library complex in \ili{Sentani} with them and staying with Rita.

Another word of thank goes to Anne-Christie Hellenthal, Francesca Moro, Victoria Nyst, and Christian Rapold. Over the years I spent many hours with them discussing “Gott und die Welt”, linguistics in general, and Papuan Malay in particular. It was always a source of great joy to me to spend time with you.

I would like to thank the LUCL office staff, namely Gea Hakker-Prins, Merel van Wijk, Anne Rose Haverkamp, and Alice Middag-Cromwijk. I would like to thank them for their assistance with the administrative questions I had. I also want to thank the Accommodate staff, namely Linda Altman, Tessa van Witzenburg, and Eric van Wijk. Their help with housing issues made living in Leiden a pleasure. Thank you all very much!

I am also very grateful to my colleagues who spent days proof-reading my original dissertation. I want to thank Sarah Hawton and Shirley McHale for their generous help. Another big thank-you goes to my colleagues Dennis Pepler, Stephen Palmstrom, Felix Kopecky and Sebastian Nordhoff for their generous help with converting the Word version of this book to LaTeX. Especially Stephen Palmstrom spent countless hours on this project - thank you so much! I am also grateful to Judy Lakeman and Jackie Gray for facilitating this help!

Finally, I want to thank Gudrun and Peter. Eure Freundschaft und Unterstützung waren ein zuverlässiger Anker mitten in den Stürmen der letzten Jahre!

My last and greatest thanks, however, goes to the triune God, who gives us life, love, and hope through his Son Jesus Christ.

\printbibliography[heading=subbibliography]
\end{refsection}

